Gene expression responses to diet quality and viral infection in Apis mellifera

### 2018-09-17 Manuscript/Text/manuscript.tex
Adam additional edits

Data visualization is a useful method to identify noise and robustness in RNA-sequencing data. In this study, we used extensive data visualization to improve the interpretation of our RNA-sequencing results. For example, the DESeq2 package comes with certain visualization options that are popular in RNA-sequencing analysis. One of these visualization is the principal component analysis (PCA) plot, which allows users to visualize the similarity between samples within a dataset. We could determine from this plot that indeed the Galbraith data may show more similarity between its replicates and differences between its treatments compared to our data (Additional file 2). However, the PCA plot only shows us information at the sample level. We wanted to investigate how these differences in the signal:to:noise ratios of the datasets would affect the structure of any resulting DEGs. As a result, we also used three plotting techniques from the bigPint package: We investigated the 1,019 virus-related DEGs from the Galbraith dataset and the 43 virus-related DEGs from our dataset using parallel coordinate lines, scatterplot matrices, and litre plots. To prevent overplotting issues in our graphics, we used a hierarchical clustering technique for the parallel coordinate lines to separate the set of DEGs into smaller groups. We also needed to examine four subsets of samples from our dataset to make effective use of the scatterplot matrices. After these tailorizations, we determined that the same patterns we saw in the PCA plots regarding the entire dataset extended down the pipeline analysis into the DEG calls: Even the DEGs from the Galbraith dataset showed more similarity between their replicates and differences between their treatments compared to those from our data. However, the 365 DEGs from the Galbraith data in Cluster 1 of Figure 3 showed an inconsistent first replicate in the treatment group (``V.1''), which was something we observed in the PCA plot. This indicates that this feature also extended down the analysis pipeline into DEG calls. Despite the differences in signal between these two datasets, there was substantial overlap in the resulting DEGs. We believe these visualization applications can be useful for future researchers analyzing RNA-sequencing data to quickly and effectively ensure that the DEG calls look reliable or at least overlap with DEG calls from similar studies that look reliable. We also expect this type of visualization exploration can be especially crucial when studying complex organisms that do not have genetic identicalness or similarity between replicates and/or when using experiments that may lack rigid design control.

One of the goals of this study was to use our RNA-sequencing data to assess whether transcriptomic responses to diet quality and virus infection provide insight into whether high quality diet can buffer bees from pathogen stress via mechanisms of ``resistance'' or ``tolerance''. Recent evidence has suggested that overall immunity is determined by more than just ``resistance'' (the reduction of pathogen fitness within the host by mechanisms of avoidance and control) \cite{resTol2}. Instead, overall immunity is related to ``resistance'' in conjunction with ``tolerance'' (the reduction of adverse effects and disease resulting from pathogens by mechanisms of healing) \cite{resTol1, resTol2}. Immune-mediated resistance and diet-driven tolerance mechanisms are costly and may compete with each other \cite{resTol1, resTol4}. Data and models have suggested that selection can favor an optimum combination of both resistance and tolerance \cite{resTol5, resTol6, resTol7, resTol8}. We attempted to address this topic through specific gene expression contrasts (Table \ref{tbl:contrasts}), accompanied by GO analysis of the associated gene lists. We found an approximately equal number of resistance (\textit{n} = 125) and tolerance (\textit{n} = 122) related candidate DEGs, suggesting both processes may be playing significant roles in dietary buffering from pathogen induced mortality. Resistance candidate DEGs had functions related to several forms of metabolism (chitin and carbohydrate), regulation of transcription, and cell adhesion. Tolerance candidate DEGs had functions related to carbohydrate metabolism and chitin metabolism; however, they also showed functions related to immune response, including RNA polymerase II transcription and regulation of response to reactive oxygen species (Figure 5A). Previous studies have shown that transcriptional pausing of RNA polymerase II may be an innate immune response in \textit{D. melanogaster} that allows for a more rapid response by increasing the accessibility of promoter regions of virally induced genes \cite{resTol11}. Moreover, circulating haemocytes in insects encapsulate and nodulate pathogens by forming a barrier between the pathogen and the host tissues. This barrier undergoes apoptosis and melanization through the phenoloxidase enzyme cascade, which produces reactive oxygen species \cite{resTol1, resTol9, resTol10}. These possible immunological defense mechanisms within our ``tolerance'' candidate DEGs and metabolic processes within our ``resistance'' candidate DEGs may provide additional evidence of feedbacks between diet and disease in honey bees \cite{AdamTothReview}.

@@

Details of the procedures we used to prepare virus inoculum, infect and feed caged honey bees, and quantify IAPV can be reviewed in our previous work \cite{adamInt, carrillo}. The statistical analysis we used to study the main and interaction effects of the two factors on mortality and IAPV titers is also described in our earlier report \cite{adamInt}.

@@

\begin{figure}[h!]
\caption{\csentence{Mortality rates for the four treatment groups, two virus groups, and two diet groups.}
Left to right: Mortality rates for the four treatment groups, two virus groups, and two diet groups. ``N'' represents non-inoculation, ``V'' represents viral inoculation, ``C'' represents Chestnut pollen, and ``R'' represents Rockrose pollen. The mortality rate data included 59 samples with 15 replicates per treatment group, except for the ``NC'' group having 14 replicates. ANOVA values and p-values for the statistical tests are listed in the text of the paper. The letters above the bars represent Tukey honest significant differences with a confidence level of 95\%.}
\end{figure}

\begin{figure}[h!]
\caption{\csentence{IAPV titer volumes for the four treatment groups, two virus groups, and two diet groups.}
Left to right: IAPV titer volumes for the four treatment groups, two virus groups, and two diet groups. ``N'' represents non-inoculation, ``V'' represents viral inoculation, ``C'' represents Chestnut pollen, and ``R'' represents Rockrose pollen. The IAPV titer data included 38 samples with 10 replicates per treatment group, except for the ``NR'' group having 8 replicates. ANOVA values and p-values for the statistical tests are listed in the text of the paper. The letters above the bars represent Tukey honest significant differences with a confidence level of 95\%.}
\end{figure}

>>>>

Data visualization is a useful method to identify noise and robustness in RNA-sequencing data \cite{edger}. In this study, we used extensive data visualization to improve the interpretation of our RNA-sequencing results. For example, the DESeq2 package comes with certain visualization options that are popular in RNA-sequencing analysis. One of these visualization is the principal component analysis (PCA) plot, which allows users to visualize the similarity between samples within a dataset. We could determine from this plot that indeed the Galbraith data may show more similarity between its replicates and differences between its treatments compared to our data (Additional file 2). However, the PCA plot only shows us information at the sample level. We wanted to investigate how these differences in the signal:to:noise ratios of the datasets would affect the structure of any resulting DEGs. As a result, we also used three plotting techniques from the bigPint package: We investigated the 1,019 virus-related DEGs from the Galbraith dataset and the 43 virus-related DEGs from our dataset using parallel coordinate lines, scatterplot matrices, and litre plots. To prevent overplotting issues in our graphics, we used a hierarchical clustering technique for the parallel coordinate lines to separate the set of DEGs into smaller groups. We also needed to examine four subsets of samples from our dataset to make effective use of the scatterplot matrices. After these tailorizations, we determined that the same patterns we saw in the PCA plots regarding the entire dataset extended down the pipeline analysis into the DEG calls: Even the DEGs from the Galbraith dataset showed more similarity between their replicates and differences between their treatments compared to those from our data. However, the 365 DEGs from the Galbraith data in Cluster 1 of Figure 3 showed an inconsistent first replicate in the treatment group (``V.1''), which was something we observed in the PCA plot. This indicates that this feature also extended down the analysis pipeline into DEG calls. Despite the differences in signal between these two datasets, there was substantial overlap in the resulting DEGs. We believe these visualization applications can be useful for future researchers analyzing RNA-sequencing data to quickly and effectively ensure that the DEG calls look reliable or at least overlap with DEG calls from similar studies that look reliable. We also expect this type of visualization exploration can be especially crucial when studying complex organisms that do not have genetic identicalness or similarity between replicates and/or when using experiments that may lack rigid design control.

One of the goals of this study was to use our RNA-sequencing data to assess whether transcriptomic responses to diet quality and virus infection provide insight into whether high quality diet can buffer bees from pathogen stress via mechanisms of ``resistance'' or ``tolerance''. Recent evidence has suggested that overall immunity is determined by more than just ``resistance'' (the reduction of pathogen fitness within the host by mechanisms of avoidance and control) \cite{resTol2}. Instead, overall immunity is related to ``resistance'' in conjunction with ``tolerance'' (the reduction of adverse effects and disease resulting from pathogens by mechanisms of healing) \cite{resTol1, resTol2}. Immune-mediated resistance and diet-driven tolerance mechanisms are costly and may compete with each other \cite{resTol1, resTol4}. Data and models have suggested that selection can favor an optimum combination of both resistance and tolerance \cite{resTol5, resTol6, resTol7, resTol8}. We attempted to address this topic through specific gene expression contrasts (Table \ref{tbl:contrasts}), accompanied by GO analysis of the associated gene lists. We found an approximately equal number of resistance (\textit{n} = 125) and tolerance (\textit{n} = 122) related candidate DEGs, suggesting both processes may be playing significant roles in dietary buffering from pathogen induced mortality. Resistance candidate DEGs had functions related to several forms of metabolism (chitin and carbohydrate), regulation of transcription, and cell adhesion. Tolerance candidate DEGs had functions related to carbohydrate metabolism and chitin metabolism; however, they also showed functions related to immune response, including RNA polymerase II transcription and regulation of response to reactive oxygen species (Figure 5A). Previous studies have shown that transcriptional pausing of RNA polymerase II may be an innate immune response in \textit{D. melanogaster} that allows for a more rapid response by increasing the accessibility of promoter regions of virally induced genes \cite{resTol11}. Moreover, circulating haemocytes in insects encapsulate and nodulate pathogens by forming a barrier between the pathogen and the host tissues. This barrier undergoes apoptosis and melanization through the phenoloxidase enzyme cascade, which produces reactive oxygen species \cite{resTol1, resTol9, resTol10}. In insects, melanization has been shown to be toxic to parasites, bacteria, fungi, and lately viruses \cite{melanization}. These possible immunological defense mechanisms within our ``tolerance'' candidate DEGs and metabolic processes within our ``resistance'' candidate DEGs may provide additional evidence of feedbacks between diet and disease in honey bees \cite{AdamTothReview}.

@@

\subsection*{Pathogen response}

Details of the procedures we used to prepare virus inoculum, infect and feed caged honey bees, and quantify IAPV can be reviewed in our previous work \cite{adamInt, carrillo}. A linear mixed effects model was used to relate the mortality rates and IAPV titers to the main and interaction effects of the diet and virus factors. The model was fitted to the data by restricted maximum likelihood (REML) using the ``lme'' function in the R package ``nlme''. Post-hoc pairwise comparisons of the four (diet and virus combination) treatment groups were performed and Benjamini-Hochberg adjusted p-values were calculated to limit familywise Type I error rates \cite{bh}.

@@

\begin{figure}[h!]
\caption{\csentence{Mortality rates for the four treatment groups, two virus groups, and two diet groups.}
Left to right: Mortality rates for the four treatment groups, two virus groups, and two diet groups. ``N'' represents non-inoculation, ``V'' represents viral inoculation, ``C'' represents Chestnut pollen, and ``R'' represents Rockrose pollen. The mortality rate data included 59 samples with 15 replicates per treatment group, except for the ``NC'' group having 14 replicates. ANOVA values and p-values for the statistical tests are listed in the text of the paper. The letters above the bars represent significant differences with a confidence level of 95\%.}
\end{figure}

\begin{figure}[h!]
\caption{\csentence{IAPV titer volumes for the four treatment groups, two virus groups, and two diet groups.}
Left to right: IAPV titer volumes for the four treatment groups, two virus groups, and two diet groups. ``N'' represents non-inoculation, ``V'' represents viral inoculation, ``C'' represents Chestnut pollen, and ``R'' represents Rockrose pollen. The IAPV titer data included 38 samples with 10 replicates per treatment group, except for the ``NR'' group having 8 replicates. ANOVA values and p-values for the statistical tests are listed in the text of the paper. The letters above the bars represent significant differences with a confidence level of 95\%.}
\end{figure}
